Do educational institutions explain SES gaps?: an interntational comparison of the SES achievement gap from 2000-2015

### 2017-09-12 paper/paper.Rnw
Finalized first draft of paper - ran everything: good

\usepackage{geometry}
 \geometry{
 a4paper,
 total={170mm,257mm},
 left=20mm,
 top=20mm,
 }
\usepackage{pdflscape}
\usepackage{amsmath}
\usepackage{subcaption}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[round, colon]{natbib}
\usepackage[colorlinks]{hyperref}
\usepackage{booktabs}
\AtBeginDocument{%
  \hypersetup{
    citecolor=blue,
    linkcolor=blue,   
    urlcolor=blue}}

\title{The not so widening achievement gap: an interntational comparison of the evolution of the SES achievement gap from 2000-2015}
\author{Jorge Cimentada}

@@

\tableofcontents

@@

\section{Literature Review}
% You have a bunch of citations to take from here
Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years. The idea of meritocracy and intergenerational transfers has motivated, for a good part of the 20th and 21st century, much of the research on social mobility and social inequalities. When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family socio-economic status (SES) and student performance are tightly linked. The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements. For a detailed review of the long literature on educational inequality, please see \citet{gamoran2001}. But fast-forward until today, we've understood the relationship much better.

Developmental psychology has been studying the subject for a long time and they find that the early stages in a child's life course are extremely important, if not the most, for cognitive development and defining personality traits \citep{duyme1999, waldfogel2006}. Economists and educators have long neglected these findings but the work of James Heckman has brought our attention to it. For example, he has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}. This means that inequality starts strong before the schooling of a child begins and schools have the potential to equalize the learning ground. These inequalities can be narrowed significantly with high quality learning environments, specially for children coming from unfavourable conditions. But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\). Although this sounds logical, it's importance is not understood entirely.

@@

For this reason, it is incomplete to study educational inequalities without considering the prevalence of intergenerational transmission of human and social capital. Social mobility has deepend our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity. The field of economics of education has also helped to understand that education is by far the investment that yields the best returns. This is not only for the child and his/her family, but to society as a whole, as it boosts economic activity, it helps the labour market improve job conditions and maintain a rapid economic growth \citep{hanushek2007}. Since \citet{coleman1966} we've spent most of our time studying the mechanisms through which this inequality comes about. Naturally, we want to do that in order to reverse it and help every child reach its fullest potential. Despite our good intentions, we've concentrated little on the magnitud of the gaps, specially in terms of cognitive abilities. We have very little evidence, in comparative terms, for which countries have big or small cognitive gaps resulting from SES origins. And even more importantly, we haven't assesed whether policy efforts to reduce inequality have actually had an impact in reducing the achievement gap over time.

However, the closest we have is the vast literature on social mobility. Thanks to these findings, we know that virtually in all countries, be it developed or developing, there is inequality of opportunity. But there is considerable variation in the magnitud of this effect. For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries \citep{esping2012, breen2007, shavit1993}.

From Denmark, for example, the literature has learned a great deal about how to improve social mobility. But that was possible by first learning that Denmark is by far one of the most mobile countries, boosting the chances of upward mobility specially for less advantaged children \citep{bjorklund2009, jaeger2007}. The important finding came when we discovered the main reason behind their social escalator: it's educational system. For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impact on improving opportunities. The education system is completely subsidized for all children, otherwise giving opportunities to children wouldn't have been able to pay. In addition to this, Denmark is recognized as a world-leader in terms of public support for its child-care system as it spends around 2\% of GDP, and has among the highest enrollment rates for children under 6 years old \citep{esping_waldfogel2012}. Moreover, it tracks students into different curricula very late compared to other Europen countries (age 16), something which has been linked to less educational inequality \citep{hanushek_woesmann_tracking}. These two traits make the Danish system uniquely effective. The early schooling experience attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable. In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.

@@

\citet{reardon2011} finds that the increase in the gaps has occured predominantely from the 1970's until the 2000's. In fact, the hard numbers suggest that the gap widened by 40-50\%. The author also estimates the rate of change using data as early as 1940 and finds an even higher increase of   75\%. Given that the studies before 1970 are less reliable in terms of comparability and sampling design, the author computes all results for both before/after 1970. To provide a definitive answer to the size of the gap, \citet{reardon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations in the year 2001. Using longitudinal data, \citet{bradbury2015} find similar results. Their empirical analysis suggests that for American 14 year olds, the gap is above 1 standard deviation but lower than 1.25. Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of  1.25-1.50 standard deviations. One important drawback of these studies is that they don't present the uncertainty of this estimates. Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy. It could be that the gap is at  1.25 standard deviations, varying up to 1.75 and down to 0.80. Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.

@@

In a follow-up study, \citet{reardon_portilla} unexpectedly uncovered a new finding: the reversal of the trend. This follow-up study concentrated solely on the kindergarten children in which they took a snapshot of the achievement gap in for 1998, 2006 and 2010. They found that the 90th/10th income gap in readiness closed modestly. Furthermore, using data from fall and spring in the same kindergarden year, they calculated that the gap narrowed at a rate of 0.01 and 0.008 standard deviations per year for mathematics and literacy between 1998 and 2010. They also calculated the same changes for a number of personality traits such as self-control and externalizing behaviour and found similar results. In contrast, \citet{reardon2011} finds that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something reasonably close to the previous estimates. Their results not only hold for the income achievement gap, but they also found a decline in the white-hispanic gap (although not for the white-black gap). It should be noted that perhaps the reversal of trend in \citet{reardon2011} would be obvious if data were available for years after 2000,  which is specifically the time that \citet{reardon_portilla} find the reversal.

The reasons why the authors find a reversal in the trend could be numerous and should be studied very closely. The authors incorporate a number of country-level indicators to explain this change and suggest that the reversal is likely due to the high increase of preschool enrollment. They build on their previous argument by suggesting that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially. Their conclusions, although suggestive, are very speculative and have no empirical support, which is why this is still an open question.

@@

\citet{anna2016_global}, building on the work of \citet{anna2016} and \citet{reardon_portilla} pooled together all the previously mentioned data, together with over 10 more studies ranging from the year 1964 until 2015 in order to discover differences between and across countries. With over 50 years of data, and over 100 countries, she finds that there seems to be a general pattern of increasing achievement gap. However, once she disentangles the relationship by country, she finds a sizable amount of heterogeneity, with some countries seeing the achievement gap narrowing, others no change at all, while others record a steady increase. This is revealing because it doesn't really pay off to look at a general average once each country has their own distinctive gap and evolution. This highlights the notion that the increasing achievement gap is clearly not universal and should be studied in context.

@@

In fact, the evolution of high/low SES gaps for preschool children might be much less marked than the same gap for high school children. The explanation, although very debated, has been gaining much support in recent years. If we remember correctly, for countries with high levels of curricular differentiation the transition from early schooling into the tracking system has been found to increase inequality of learning \citep{hanushek_woesmann_tracking}. Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs \citep{werfhorst_mijs}. Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts (because there have been tracking reforms in many countries, introducing as well as eliminating tracking structures) nor the same between ages, because tracking/no tracking might exarcebate the achievement gap.

@@

More formally, we're interested in studying
a) Which countries are experiencing changes in the achievement gap;
b) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind;
c) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;
d) What explains this cross-country evolution;

@@

\section{Hypothesis}

@@

\begin{itemize}
\item Hypothesis 1: In countries where there is a high degree of tracking we should expect the top and bottom to be evolving at the rate given that tracking is thought to maximize the learning experience of both groups. For countries with low levels of tracking we should expect the top and bottom groups to be closer to each other and for the gap to be closing since the both groups will be in the same track (this doesn't mean that the top is loweing their score but that the bottom is most likely catching up).

@@

To investigate the above mentioned questions I will be using the Programme for International Student Assessment (PISA). PISA is a survey carried out every three years that aims to evaluate education systems by testing the skills and knowledge of 15-year-old students. Currrently, PISA has six waves starting in 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 developed/developing countries.

@@

Aside from SES, the other most relevant variables are test scores for mathematics and literacy. PISA does not provide a single test result for each respondent. Instead, it provides a \emph{series} of 'plausible values' that the child could actually score. As explained in the PISA manual \citep{pisa2012_technical}, these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured (the true distribution of the possible scores a student can achieve).

@@

The aim of this paper is to identify, disaggregate and explain country trends in the achievement gap for several countries. To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps \citep{fryer2004, hanushek_woesmann_tracking, saw2016, bradbury2015, byun2010}. The actual calculation of the achievement gap varies substantially and different strategies have been implemented. For example, \citet{micklewright} calculates the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution. Although in principle you should be able to capture some type of SES effect like this, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education or some other SES proxy. \citet{saw2016}, for instance, used parental education as a proxy of SES, whereas \citet{byun2010} use a similar SES index as ours, but created by them.

@@

for each wave, where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (following the same strategy as \citet{reardon2011} \footnote{This does not mess up our analysis by masking age-specific gaps because all students in the sample are 15 year olds. Control for aging is simply to control for monthly differences in ages.}) weighted by the student sample weights.

@@

where \begin{math}r\end{math} is the reliability score of the wave \footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}. The realibility estimates are calculated using Item Response Theory (IRT) analogues of traditional estimates of person separation reliability such as internal consistency. Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own. At the moment of writing this paper, they were unavailable. For this reason we implement the analysis following the original work of \citet{reardon2011}}. Each PISA survey provides a reliability indicator which we use accordingly. This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.

@@

Finally, we calculate the SES achievement gap for each country by extracting the fitted \(\alpha\) and \(\beta\) for each country \(j\) and calculating the difference of the predicted score of high and low SES.

@@

The first table shows a description of the sample size and mean score of both top and bottom SES groups for only the first and last time point. One main concern from the planned analysis is that getting the top 90th percentile and bottom 10th percentile would result in a small sample size. The table suggests that we have a reasonable number of respondents to actually estimate gaps accurately. Moreover, we can see that in all instances the bottom SES group has a lower score than all top SES groups.

We see that in some countries like Finland and Sweden the average low SES scores is actually above the average score of 0, suggesting they're very equal countries. However, most countries, like Spain, Poland, Hungary and Germany, their average low SES scores are much lower than the country average, which is 0. We also see some countries with major changes from year 2000 to 2015, with, for example, Australia decreasing the low SES gap from 0.22 standard deviations to -0.15, well below the country average. In the next section we take a look at this in a more detailed fashion, considering all years in between.

@@

On the other hand, we do have some countries which show a decrease in the SES achievement gap. Poland decreased by about \Sexpr{diff_data$math[["Poland"]]} and Denmark by \Sexpr{diff_data$math[["Denmark"]]}. However, the most notable cases are the United States and Germany. These two countries show a very high level of dispersion in the year 2000 with SES gaps of over 2 standard deviations. But in the 15-year time trend both countries reduced the gaps by \Sexpr{diff_data$math[["United States"]]} and \Sexpr{diff_data$math[["Germany"]]} respectively. Their distinctively big gaps in 2000 also show up in the work of \citet{micklewright}. This corroborates the findings of \citet{reardon_portilla}, which found a decreasing gap for kindergarteners. This decline is evident also for 15 year olds, suggesting it might be more of an institutional change rather than a specific grade-level policy. Do note that the trendline is for both mathematics and literacy. If it were only for mathematics, then the trendline would probably be even more extreme. Despite this, pooling both subjects gives even more strength to the analysis given that the results stand for both subjects together \footnote{Same graph is availabe for reading upon request, where we find pretty much the same results if not even more steep/downward slopes.}. We also plot the 80/20 and 70/30 SES gaps and find that very similar patterns as in the 90/10 SES gap (see figure 10 in the appendix).

@@

We find that the initial gap for the U.S in 2000 is \Sexpr{round(us_math_2000, 2)} varying between \Sexpr{round(us_math_2000_lower, 2)} and \Sexpr{round(us_math_2000_upper, 2)}. This gap is much higher than found in previous studies, but again, we don't know how much previous estimates vary and nobody has really studied the 15-year old achievement gap precisely. All country-level studies pool many studies together and control for age, removing the age-specific gap effect.

Figure 3 below takes a more direct approach and looks at the percentage change from the first and last timepoint available. Each data point has been computed together with it's 50\% and 95\% uncertainty interval \footnote{Each of these uncertainty intervals were computed using a 500-replicate bootstrap}.

@@

Generally speaking, we see that most countries increased their achievement gap over time. France had an average increase of about 80\% since 2000 varying down to 40\%, whereas Germany had a similar figure but decreasing. Many of the countries we saw before that didn't have a very steep slope, such as Hungary or Australia, in fact had increases of about 40\% of their gap. In contrast, we see that the U.S and Poland had also quite significant decreases of about 40\%. The benefit of presenting these estimates this way is that we can actually asses the uncertainty of each calculation, and we do find that some of these estimates have wide variability. Despite this, most countries show a clear sign of either decreasing or increasing. We plot the same graph for the 80/20 and 70/30 SES gaps in the appendix and the results hold for these achievement gaps as well.

@@

Something reassuring is that mathematics and reading (not presented) follow basically the same trend across all countries \footnote{Reading had an even bigger decrease for similar countries, and a much smaller increase. The previos two graphs for reading are available upon request}. This means that the result is not an artifact of chance. However, it's important to disentangle where is this gap originating from. Is this because the top are improving while the bottom decreases? Or is it that the bottom is catching up? Up next we plot the same graph but show the divergent patterns between high/low SES origins.

@@

Figure 4 shows the evolution for both low and high SES groups separately. The middle line is a summary line showing whether one of the groups is growing/shrinking faster than the other. We find plenty of variation in these patterns. For example, in the United States the top seems to be equalizing much stronger than than the bottom is catching up. The U.K seems to be following the same path as the United States as well. On the other hand, in Poland the gap seems to be closing because the bottom SES group is catching up much faster. The Netherlands shows a similar pattern as the U.S, U.K patterns but upon closer inspection the explanation is very different. We see that the slope is negative (like in the previous two countries) but that's because the low SES group is decreasing at a greater rate than the high SES is decreasing.

We now pay attention to the flat summary lines. These lines should be interpreted with caution because it doesn't mean that the gap is not increasing. Denmark, Belgium, Netherlands, Norway and Spain show a flat line because the gap is growing very little, if at all. On the other hand, Sweden, France, Finland, New Zealand, Austria, among other countries, show a flat line because both groups are distancing themselves at a very similar rate. These results highlight the importance of not only summarizing average achievement gaps. The source of these gaps vary greatly between countries and we see how each of these patterns contibutes to the overall inequality of a country.

After analyzing the trend for each country and where this trend is coming from we start to find some interesting results. We see that the United States is closing the achievement gap at a very rapid rate, but at the same time we notice that it's because the top SES group is going down faster than the bottom is catching up. In contrast, we see countries like Germany and Poland where the gap is closing as well but because bottom group is catching up faster than the high SES group is coming down. In fact, the Polish case is the most egalitarian of all the countries presented here; the bottom group is catching up at a rapid pace and the top group is maintaining their high levels of performance, fostering a trully equalizing effect.

@@

\subsection{The role of curricular tracking in explaining inequalities}

@@

We now move to the final section of the paper, where we discuss what is a likely explanation to the patterns from before. The reason why countries differ in their evolution of inequality is still unknown. There is ample evidence showing the inequality between countries, to a certain extent, can be explained by the degree of tracking \citep{hanushek_woesmann_tracking}. But we are also interested in what explains the between country evolution of inequality, that is, the explanation as to why in certain countries it is increasing more than others over time.

>>>>

\usepackage{geometry}
 \geometry{
 a4paper,
 total={170mm,257mm},
 left=20mm,
 top=20mm,
 }
\usepackage{pdflscape}
\usepackage{amsmath}
\usepackage{subcaption}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{titling}
\usepackage[round, colon]{natbib}
\usepackage[colorlinks]{hyperref}
\usepackage{booktabs}
\AtBeginDocument{%
  \hypersetup{
    citecolor=blue,
    linkcolor=blue,   
    urlcolor=blue}}

\title{Do educational institutions explain SES gaps?: an interntational comparison of the SES achievement gap from 2000-2015}
\author{Jorge Cimentada}

@@

\begin{abstract}
The literature on achievement inequality has only recently started studying the evolution of the socio-economic achievement gap in cognitive abilities and it has unconvered a wide variety of results. The main finding comes from American research and it shows that the gap has been widening by about 40-50\% in the last 30 year. Other international comparisons show that there's important variation among developed countries, but the research on this area is scarce. This paper looks to study whether there are specific patterns in the evolution of the gap in a comparative perspective. We find that there is considerable variation in the way in which the gap is evolving, with the U.S and Germany closing at a rapid pace and France widening at a similar rate. We find that this gap is strongly correlated with a countrys overall performance and simulations show that switching to less tracking can help decrease a countries SES gap by about 20\%. We also find that tracking and vocational enrollment explain over one third of the evolution of the achievement gap and show that the relationship is conditioned by a strong interaction: low tracking is associated with a small achievement gap, but high levels of tracking coped with high vocational enrollment can also remedy the adversive effects of tracking.
\end{abstract}

@@

% You have a bunch of citations to take from here
Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years. The idea of meritocracy and intergenerational transfers has motivated, for a good part of the 20th and 21st century, much of the research on social mobility and social inequalities. When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family socio-economic status (SES) and student performance are tightly linked. The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements \footnote{For a detailed review of the long literature on educational inequality, please see \citet{gamoran2001}.}. But fast-forward until today, we've understood the relationship much better.

Developmental psychology has been studying the subject for a long time and they find that the early stages in a child's life course are extremely important, if not the most, for cognitive development and defining personality traits \citep{duyme1999, waldfogel2006}. Economists and educators have long neglected these findings until the work of James Heckman brought our attention to it. For example, he has showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}. This means that inequality starts strong before the schooling of a child begins and schools have the potential to equalize the learning ground. These inequalities can be narrowed significantly with high quality learning environments, specially for children coming from unfavourable conditions. But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiencies of time \(t-1\). Although this sounds logical, it's importance is not understood entirely.

@@

For this reason, it is incomplete to study educational inequalities without considering the prevalence of intergenerational transmission of human and social capital. Social mobility has deepend our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity. The field of economics of education has also helped to understand that early high-quality education is by far the best investment that yields the highest returns. This is not only for the child and his/her family, but to society as a whole as it boosts economic activity, it helps the labour market improve job conditions and maintain a rapid economic growth \citep{hanushek2007}. Since \citet{coleman1966} we've spent most of our time studying the mechanisms through which this inequality comes about. Naturally, we want to do that in order to reverse it and help every child reach its fullest potential. Despite our good intentions, we've concentrated little on the magnitud of the gaps, specially in terms of cognitive abilities. We have very little comparative evidence on which countries have big or small cognitive gaps resulting from SES origins. And even more importantly, we haven't assesed whether policy efforts to reduce inequality have actually had an impact in reducing the achievement gap over time.

Despite these short-comings, we have indirect evidence of this coming from the vast literature on social mobility. Thanks to these findings, we know that virtually in all countries, be it developed or developing, there is inequality of opportunity. But there is considerable variation in the magnitud of this effect. For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries \citep{esping2012, breen2007, shavit1993}.

From Denmark, for example, the literature has learned a great deal about how to improve social mobility. But that was possible by first learning that Denmark is by far one of the most mobile countries; this is the case specially for families of low-ses origin \citep{bjorklund2009, jaeger2007}. The important finding came when we discovered the main reason behind their social escalator: it's educational system. For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impact on improving opportunities. The education system is completely subsidized for all children, otherwise giving opportunities to families who wouldn't have been able to pay. In addition to this, Denmark is recognized as a world-leader in terms of public support for its child-care system as it spends around 2\% of GDP, and has among the highest enrollment rates for children under 6 years old \citep{esping_waldfogel2012}. Moreover, it separates students into different tracks very late compared to other Europen countries (age 16), something which has been linked to less educational inequality \citep{hanushek_woesmann_tracking}. These two traits make the Danish system uniquely effective. The early schooling experience attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable. In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.

@@

\citet{reardon2011} finds that the increase in the gaps has occured predominantely from the 1970's until the 2000's. In fact, the hard numbers suggest that the gap widened by 40-50\%. The author also estimates the rate of change using data as early as 1940 and finds an even higher increase of   75\%. Given that the studies before 1970 are less reliable in terms of comparability and sampling design, the author computes all results for both before/after 1970. To provide a definitive answer to the size of the gap, \citet{reardon2011} concludes that the 90/10 income gap in the U.S is at about 1.25 standard deviations in the year 2001. Using longitudinal data, \citet{bradbury2015} find similar results. Their empirical analysis suggests that for American 14 year olds, the gap is above 1 standard deviation but lower than 1.25. Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of  1.25-1.50 standard deviations. One important drawback of these studies is that they don't present the uncertainty of these estimates. Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy. It could be that the gap is at  1.25 standard deviations, varying up to 1.75 and down to 0.80. Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.

@@

In a follow-up study, \citet{reardon_portilla} unexpectedly uncovered a new finding: the reversal of the trend. This follow-up study concentrated solely on kindergarten children for the years 1998, 2006 and 2010. They found that the 90th/10th income gap in readiness closed modestly. Furthermore, using data from fall and spring in the same kindergarden year, they calculated that the gap narrowed at a rate of 0.01 and 0.008 standard deviations per year for mathematics and literacy between 1998 and 2010. They also calculated the same changes for a number of personality traits such as self-control and externalizing behaviour and found similar results. In contrast, \citet{reardon2011} finds that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something reasonably close to the previous estimates. Their results not only hold for the income achievement gap, but they also found a decline in the white-hispanic gap (although not for the white-black gap). It should be noted that perhaps the reversal of the trend in \citet{reardon2011} would had been evident if data were available for years after 2000, the time-point from which \citet{reardon_portilla} find the reversal.

The reasons why the authors find a reversal in the trend could be numerous and should be studied very closely. They incorporate a number of country-level indicators to explain this change and suggest that the reversal is likely due to the high increase of preschool enrollment. They build on their previous argument by suggesting that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially. Their conclusions, although suggestive, are very speculative and have no empirical support, which is why this is still an open question.

@@

\citet{anna2016_global}, building on the work of \citet{anna2016} and \citet{reardon_portilla} pooled together all the previously mentioned data, together with over 10 more studies ranging from 1964 to 2015 in order to discover differences between and across countries. With over 50 years of data, and over 100 countries, she finds that there seems to be a general pattern of increasing achievement gap. However, once she disentangles the relationship by country, she finds a sizable amount of heterogeneity, with some countries seeing the achievement gap narrowing, others no change at all, while others record a steady increase. This is revealing because it doesn't really pay off to look at a general average once each country has their own distinctive gap and evolution. This highlights the notion that the increasing achievement gap is clearly not universal and should be studied in context.

@@

In fact, the evolution of high/low SES gaps for preschool children might be much less marked than the same gap for high school children. The explanation, although very debated, has been gaining much support in recent years. For countries with high levels of curricular differentiation the transition from early schooling into the tracking system has been found to increase inequality of learning \citep{hanushek_woesmann_tracking}. Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs \citep{werfhorst_mijs}. Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts nor the same between ages, because tracking/no tracking might exarcebate the achievement gap.

@@

More formally, we're interested in studying
a) Which countries are experiencing changes in the achievement gap;
b) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind;
c) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;
d) What explains this cross-country evolution.

@@

\subsection{Hypothesis}

@@

\begin{itemize}
\item Hypothesis 1: In countries where there is a high degree of tracking we should expect the top and bottom to be evolving at a similar rate given that tracking is thought to maximize the learning experience of both groups. For countries with low levels of tracking we should expect the top and bottom groups to be closer to each other and for the gap to be closing since the both groups will be in the same track (this doesn't mean that the top is lowering their score but that the bottom is most likely catching up).

@@

To investigate the above mentioned questions we will be using the Programme for International Student Assessment (PISA). PISA is a survey carried out every three years that aims to evaluate education systems by testing the skills and knowledge of 15-year-old students. Currrently, PISA has six waves starting in 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 developed/developing countries.

@@

Aside from SES, the other relevant variables are test scores for mathematics and literacy. PISA does not provide a single test result for each respondent. Instead, it provides a \emph{series} of 'plausible values' that the child could actually score. As explained in the PISA manual \citep{pisa2012_technical}, these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured (the true distribution of the possible scores a student can achieve).

@@

The aim of this paper is to identify, disaggregate and explain country trends in the achievement gap for several countries. To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps \citep{fryer2004, hanushek_woesmann_tracking, saw2016, bradbury2015, byun2010}. The actual calculation of the achievement gap varies substantially and different strategies have been implemented. For example, \citet{micklewright} calculates the difference in achievement by crudely subtracting the difference between the 95th and 5th percentile of the mathematics distribution. Although in principle you should be able to capture some type of SES effect like this, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education or some other SES proxy. \citet{saw2016}, for instance, used parental education as a proxy of SES, whereas \citet{byun2010} use a similar SES index as ours, but created by them.

@@

for each wave, where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (following the same strategy as \citet{reardon2011} \footnote{This does not mess up our analysis by masking age-specific gaps because all students in the sample are 15 year olds. Controling for age is simply to adjust for monthly differences in ages.}) weighted by the student sample weights.

@@

where \begin{math}r\end{math} is the reliability score of the PISA wave \footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}. The realibility estimates are calculated using Item Response Theory (IRT) analogues of traditional estimates of person separation reliability such as internal consistency. Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own. At the moment of writing this paper, they were unavailable. For these reasons we implement the analysis following the original work of \citet{reardon2011}}. Each PISA survey provides a reliability indicator which we use accordingly. This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.

@@

Finally, we calculate the SES achievement gap for each country by extracting the fitted \(\alpha\) and \(\beta\) for each country \(j\) and calculating the difference of the predicted score of High and Low SES.

@@

Table 1 shows a description of the sample size and mean score of both top and bottom SES groups for only the first and last time point. One main concern from the planned analysis is that getting the top 90th percentile and bottom 10th percentile would result in a small sample size. The table suggests that we have a reasonable number of respondents to actually estimate gaps accurately. Moreover, we can see that in all instances the bottom SES group has a lower score than all top SES groups.

We see that in some countries like Finland and Sweden the average low SES scores is actually above the average score of 0, suggesting they're very equal countries. However, most countries, like Spain, Poland, Hungary and Germany, their average low SES scores are much lower than the country average. We also see some countries with major changes from year 2000 to 2015, with, for example, Australia decreasing the low SES gap from 0.22 standard deviations to -0.15, well below the country average. In the next section we take a look at this in a more detailed fashion, considering all years in between.

@@

On the other hand, we do have some countries which show a decrease in the SES achievement gap. Poland decreased by about \Sexpr{diff_data$math[["Poland"]]} and Denmark by \Sexpr{diff_data$math[["Denmark"]]}. However, the most notable cases are the United States and Germany. These two countries show a very high level of dispersion in the year 2000 with SES gaps of over 2 standard deviations. But in the 15-year time trend both countries reduced the gaps by \Sexpr{diff_data$math[["United States"]]} and \Sexpr{diff_data$math[["Germany"]]} respectively. Their distinctively big gaps in 2000 also show up in the work of \citet{micklewright}. This corroborates the findings of \citet{reardon_portilla}, which found a decreasing gap for kindergarteners. This decline is evident also for 15 year olds, suggesting it might be more of an institutional change rather than a specific grade-level policy. Do note that the trendline is for both mathematics and literacy. If it were only for mathematics, then the trendline would probably be even more pronounced. Despite this, pooling both subjects gives even more strength to the analysis given that the results stand for both subjects together \footnote{Same graph is availabe for reading upon request, where we find pretty much the same results if not even more steep/downward slopes.}. We also plot the 80/20 and 70/30 SES gaps and find very similar patterns (see figure 13 in the appendix).

@@

We find that the initial gap for the U.S in 2000 is \Sexpr{round(us_math_2000, 2)} varying between \Sexpr{round(us_math_2000_lower, 2)} and \Sexpr{round(us_math_2000_upper, 2)}. This gap is much higher than found in previous studies, but again, we don't know how much previous estimates vary and nobody has really studied the 15-year old achievement gap in isolation. All country-level studies pool many studies together and control for age, removing the age-specific gap effect.

Figure 3 takes a more direct approach and looks at the percentage change from the first and last timepoint available. Each data point has been computed together with it's 50\% and 95\% uncertainty interval \footnote{Each of these uncertainty intervals were computed using a 500-replicate bootstrap}.

@@

Generally speaking, we see that most countries increased their achievement gap over time. France had an average increase of about 80\% since 2000 varying down to 40\%, whereas Germany had a similar figure but decreasing. Many of the countries we saw before that didn't have a very steep slope, such as Hungary or Australia, in fact had increases of about 40\% of their gap. In contrast, we see that the U.S and Poland had also quite significant decreases of about 40\%. The benefit of presenting these estimates this way is that we can actually asses the uncertainty of each calculation, and we do find that they have wide variability. Despite this, most countries show a clear sign of either decreasing or increasing. We plot the same graph for the 80/20 (figure 14) and 70/30 (figure 15) SES gaps in the appendix and the results hold for these achievement gaps as well.

@@

Something reassuring is that mathematics and reading (not presented) follow basically the same trend across all countries \footnote{Reading had an even bigger decrease for similar countries, and a much smaller increase. The previos two graphs for reading are available upon request}. This means that the result is not an artifact of chance alone. However, it's important to disentangle where is this gap originating from. Is this because the top are improving while the bottom decreases? Or is it that the bottom is catching up? Up next we plot the same graph but show the divergent patterns between High/Low SES origins.

@@

Figure 4 shows the evolution for both low and high SES groups separately. The middle line is a summary line showing whether one of the groups is growing/shrinking faster than the other. We find plenty of variation in these patterns. For example, in the United States the top seems to be equalizing much stronger than the bottom is catching up. The U.K seems to be following the same path as the United States as well. On the other hand, in Poland the gap seems to be closing because the bottom SES group is catching up much faster. The Netherlands shows a similar pattern as the U.S and U.K patterns but upon closer inspection the explanation is very different. We see that the slope is negative (like in the previous two countries) but that's because the low SES group is decreasing at a greater rate than the high SES is decreasing.

We now pay attention to the flat summary lines. These lines should be interpreted with caution because it doesn't mean that the gap is not changing. Denmark, Belgium, Netherlands, Norway and Spain show a flat line because the gap is growing very little, if at all. On the other hand, Sweden, France, Finland, New Zealand, Austria, among other countries, show a flat line because both groups are distancing themselves at a very similar rate. These results highlight the importance of not only summarizing average achievement gaps: the source of the gap varies greatly between countries and we see how each of these patterns contibutes to the overall inequality of a country.

After analyzing the trend of each country and where it's coming from, we start to find some interesting results. We see that the United States is closing the achievement gap at a very rapid rate, but at the same time we notice that it's because the top SES group is going down faster than the bottom is catching up. In contrast, we see countries like Germany and Poland where the gap is closing as well but because the bottom group is catching up faster than the high SES group is coming down. In fact, the Polish case is the most egalitarian of all: the bottom group is catching up at a rapid pace and the top group is maintaining their high levels of performance, fostering a trully equalizing effect.

@@

\subsection{The role of curricular tracking in explaining the achievement gap}

@@

We now move to the final section of the paper, where we discuss what is a likely explanation to these patterns. The reason why countries differ in their evolution of inequality is still unknown. There is ample evidence showing the inequality between countries, to a certain extent, can be explained by the degree of tracking \citep{hanushek_woesmann_tracking}. But we are also interested in what explains the between country evolution of inequality, that is, the explanation as to why in certain countries it is increasing more than others over time.
